---
title: "Tour de France plots"
output:
  html_document:
    df_print: paged
---

# SETUP
Load libraries.
```{r setup}
library(tidyverse)
library(rvest)
library(maps)
library(mapproj)
library(countrycode)
library(plotly)
library(lubridate)
```

Import data.
```{r import_data}
tour_data <- read_csv("data/tour_data.csv")

spec(tour_data)
```

# WRANGLE DATA
Rearrange data for plotting (one row per location).
```{r tour_data}
# Create year variable from date.
tour_data <- tour_data %>%
  mutate(year = year(date))

# Create date variable date_1900 with all dates in same year, 1900 arbitrarily. 
tour_data <- tour_data %>%
  mutate(date_1900 = make_date(year = 1900L, month = month(date), day = day(date)))

# Create variable tour_day to be day of tour, and section_number to be section of tour (ie if there are more than one stage or substage on a day they are given a different sectionnumber but the same tour_day).
tour_data <- tour_data %>%
  arrange(date) %>%
  group_by(year) %>%
  mutate(tour_day = lubridate::time_length(interval(min(date), date), "day"),
         counter = if_else(terrain_group == "Rest day", 0, 1),
         section_number = if_else(terrain_group == "Rest day", -1, cumsum(counter))) %>%
  select(-counter) %>%
  ungroup()

tour_start_loc <- tour_data %>%
  select(-end_lat,
         -end_long,
         -end_place,
         -end_country,
         -end_gnid,
         -end_elev) %>%
  rename(
    place = "start_place",
    country = "start_country",
    lat = "start_lat",
    long = "start_long",
    elev = "start_elev",
    gnid = "start_gnid"
  ) %>%
  mutate(position = "start")

tour_end_loc <- tour_data %>%
  select(-start_lat,
         -start_long,
         -start_place,
         -start_country,
         -start_gnid,
         -start_elev) %>%
  rename(
    place = "end_place",
    country = "end_country",
    lat = "end_lat",
    long = "end_long",
    elev = "end_elev",
    gnid = "end_gnid"
  ) %>%
  mutate(position = "end")

tour_all_loc <- dplyr::bind_rows(tour_start_loc, tour_end_loc) %>%
  arrange(date, position)

rm(tour_start_loc, tour_end_loc)

tour_all_loc[1:20, ]
```

# TILE PLOT OF STAGES
Plot each year as a bar, with colours for each terrain_group.
```{r plot_stages}
stage_type_plot <- tour_data %>%
  filter(terrain_group != "Rest day") %>%
  arrange(date) %>%
  group_by(year) %>%
  mutate(counter = 1,
         section_number = cumsum(counter)) %>%
  select(-counter) %>%
  ungroup() %>%
  ggplot(aes(x = year, y  = section_number, fill = terrain_group)) +
  geom_tile (colour = "#000000") +
  scale_fill_manual(values=c("#008B00", "#0A5CF5", "#FFD700", "#CD0000", "#000000","#9400D3", "#0000EE")) + 
  scale_x_continuous(breaks=c(1903,1914, 1939, 1950, 1970, 1990, 2010, 2019)) +
  theme(panel.background = element_rect(fill = "white")) + theme(plot.subtitle = element_text(vjust = 1), 
    panel.grid.major = element_line(linetype = "blank"), 
    panel.grid.minor = element_line(linetype = "blank"), 
    axis.title = element_text(size = 80), 
    axis.text = element_text(size = 80, colour = "gray16"), 
    axis.text.x = element_text(size = 80, 
        colour = "gray16"), axis.text.y = element_text(size = 80), 
    plot.title = element_text(size = 100), 
    legend.text = element_text(size = 80), 
    legend.title = element_text(size = 80)) +labs(title = "Stages of the Tour", x = "Year", y = "Stage", fill = "Type of stage") + 
    theme(legend.key = element_rect(size = 20)) + 
    theme(axis.title.x = element_text(vjust = 15), axis.text.x = element_text(vjust = 20),
    plot.title = element_text(vjust = -20))

ggsave("tour_stage_type.pdf", stage_type_plot, width = 1485, height = 1000, units = "mm", limitsize = FALSE)
```


# PLOT COUNTRIES
Create plot of countries Tour has gone through since the first Tour to the latest.
```{r plot_borders}
world_map <- map_data(map = "world")

countries <-
  c("Belgium",
    "Germany",
    "Ireland",
    "Italy",
    "Luxembourg",
    "Netherlands",
    "Spain",
    "Switzerland",
    "UK")

borders_france <- map_data(map = "world", region = "France")
borders_others <- map_data(map = "world", region = countries)

# Plot borders of countries
borders_plot <- ggplot() +
  geom_polygon(data = borders_france, aes(x = long, y = lat, group = group), fill = "#FEED00", colour = "#FDFEFE") +
  geom_polygon(data = borders_others, aes(x = long, y = lat, group = group), fill = "#F2F3F4", colour = "#FDFEFE") +
  coord_map() +
  theme_void()

borders_plot
```

# PLOT TOUR LOCATIONS
Plot locations of Tour stages.
```{r plot_loc}
# Plot all locations of tour stages
locations_plot <- borders_plot +
  geom_point(data = tour_all_loc, aes(x = long, y = lat), size = 0.25, colour = "black")

locations_plot

locations_bin <- ggplot(data = tour_all_loc, aes(x = long, y = lat)) +
  stat_binhex(bins = 15, colour = "white", na.rm = TRUE) +
  scale_fill_gradientn(colours = c("#FEED00", "#262525"), name = "Frequency", na.value = NA)

locations_bin
```

# PLOT STAGES BY TERRAIN TYPE
Plot locations by terrain on map of Europe.
```{r plot_stage_type_loc}
# Split stages by stage type (stage, sub_stage and prologue)
loc_terrain<- borders_plot +
  geom_point(data = tour_all_loc, aes(x = long, y = lat), size = 0.25) +
  facet_wrap(~ terrain_group)

loc_terrain
```

Plot stages by elevation change, average elevation and colour by terrain_group.
```{r plot_terrain_elev}
tour_data <- tour_data %>%
  mutate(elev_change = end_elev - start_elev,
         elev_mean = (start_elev + end_elev)/2)

# Scatter plot
ggplot(tour_data, aes(elev_change, end_elev)) +
  geom_point() +
  scale_colour_manual(values=c("#008B00", "#FFD700","#0000EE", "#CD0000", "#000000","#9400D3", "#00B2EE")) + 
  theme(panel.background = element_rect(fill = "white")) +
  facet_wrap(~terrain_group)

ggplot(tour_data, aes(elev_change, end_elev)) +
  geom_bin2d() +
  scale_colour_manual(values=c("#008B00", "#FFD700","#0000EE", "#CD0000", "#000000","#9400D3", "#00B2EE")) + 
  theme(panel.background = element_rect(fill = "white")) +
  facet_wrap(~terrain_group)

# Hilly only 
tour_data %>%
  filter(terrain_group == "Hilly") %>%
ggplot(aes(elev_change, end_elev)) +
  geom_bin2d() +
  scale_colour_manual(values=c("#008B00", "#FFD700","#0000EE", "#CD0000", "#000000","#9400D3", "#00B2EE")) + 
  theme(panel.background = element_rect(fill = "white")) +
  facet_wrap(~terrain_group)

# Mountain only 
tour_data %>%
  filter(terrain_group == "Mountain") %>%
ggplot(aes(elev_change, end_elev)) +
  geom_bin2d() +
  scale_colour_manual(values=c("#008B00", "#FFD700","#0000EE", "#CD0000", "#000000","#9400D3", "#00B2EE")) + 
  theme(panel.background = element_rect(fill = "white")) +
  facet_wrap(~terrain_group)
```

# PLOT TOUR WINNERS
```{r plot_stage_winners}
winner_plot <- tour_data %>%
  filter(terrain_group != "Rest day", year != 2019) %>%
  mutate(winner_name_short = if_else(is.na(winner_country), winner_name, str_replace(winner_name, "^[A-Z].* ", "")),
         winner_label = paste0(winner_name_short, if_else(is.na(winner_country), "",  paste0(" (", winner_country, ")")))) %>%
  ggplot(aes(section_number, distance_km, label = winner_label)) +
  geom_text(hjust = -0.05,
            nudge_x = 0,
            angle = 90) +
  geom_bar(stat = "identity", aes(fill = terrain_group)) +
    theme(panel.background = element_rect(fill = "white")) +
    facet_wrap( ~ year, ncol = 6) +
  ylim(0, 900) +
  scale_fill_manual(values=c("#008B00", "#0A5CF5", "#FFD700", "#CD0000", "#000000","#9400D3", "#0000EE")) +
  theme(
    plot.title = element_text(
      size = 50,
      face = "bold",
      hjust = 0.5,
      margin = margin(t = 10, b = 40, unit = "pt")
    ),
    strip.text = element_text(size = 25, face = "bold", colour = "black"),
    axis.text.y = element_text(size = 25),
    axis.title.y = element_text(size = 40),
    axis.title.x = element_text(size = 40),
    axis.text.x = element_text(size = 0),
    legend.title = element_text(size = 40), 
    legend.text = element_text(size = 35),
    strip.background = element_rect(fill = "white"),
    axis.ticks.x = element_blank(), 
   ) +
  labs(title = "Stage winners of the Tour de France", x = "Stage", y = "Distance cycled (km)", fill = "Stage type")

ggsave("stage_winners.pdf", winner_plot, width = 1000, height = 1485, units = "mm", limitsize = FALSE)
```

# PLOT ELEVATION OF TOURS
Categorise locations as being in Alps, Pyrenees or other.
```{r mount_range}
# Assign location to Alps or Pyrenees based on rectangular box with GPS coordinates as cartesian.
tour_all_loc <- tour_all_loc %>%
  mutate(mountain_range = case_when(
    lat > 43.28 & lat < 47.38 & long > 5.24 & long < 7.60 ~ "Alps",
    lat > 41.98 & lat < 43.11 & long > -1.69 & long < 2.78 ~ "Pyrenees",
    lat > 44.93 & lat < 45.92 & long > 2.41 & long < 3.08 ~ "Massif Central",
    TRUE ~ "Other"
  ))

# Plot locations and colour by mountain_range to check worked.
borders_france <- map_data(map = "world", region = "France")

cols = c("Pyrenees" = "green", "Massif Central" = "red", "Alps" = "blue", "Other" = "black")

mount_plot <- ggplot() +
  geom_polygon(data = borders_france, aes(x = long, y = lat, group = group), fill = "#FEED00", colour = "#FDFEFE") +
  coord_map() +
  theme_void() +
  geom_point(data = tour_all_loc, aes(long, lat, colour = mountain_range)) +
  scale_color_manual(values = cols)

mount_plot
```


```{r plot_elevation}
tour_elevation <- tour_all_loc %>%
  group_by(year) %>%
  mutate(tour_section_progress = section_number / max(section_number),
         end_x = lead(tour_section_progress),
         end_y = lead(elev)) %>%
  ungroup() %>%
  ggplot() +
  geom_segment(aes(x = tour_section_progress, y = elev, xend = end_x, yend = end_y, colour = mountain_range), size = 2) +
  theme(panel.background = element_rect(fill = "white")) +
  facet_wrap( ~ year, ncol = 10) +
  scale_color_manual(values = cols, breaks = c("Alps", "Pyrenees", "Massif Central", "Other")) +
  theme(
    plot.title = element_text(
      size = 50,
      face = "bold",
      hjust = 0.5,
      margin = margin(t = 10, b = 40, unit = "pt")
    ),
    strip.text = element_text(size = 25, face = "bold"),
    axis.text.y = element_text(size = 25),
    axis.title.y = element_text(size = 40),
    axis.title.x = element_text(size = 40),
    axis.text.x = element_text(size = 0),
    legend.title = element_text(size = 40), 
    legend.text = element_text(size = 35),
    # strip.background = element_rect(fill = "#FEED00"),
    axis.ticks.x = element_blank(), 
   ) +
  labs(title = "L'altitude des Tours de France", x = "Tour % progress by stage", y = "L'altitude (metres)", colour = "Mountain range")

ggsave("tour_elevation.pdf", tour_elevation, width = 1000, height = 1485, units = "mm", limitsize = FALSE)
```

# PLOT DISTANCE CYCLED IN EACH STAGE
```{r plot_dist_cyc}
total_dist <- tour_data %>%
  group_by(year) %>%
  summarise(dist_cyc_tot_km = sum(distance_km), dist_crow_tot_km = sum(distance_crow_km)) %>%
  ungroup()

tour_dist <- tour_data %>%
  filter(terrain_group != "Rest day") %>%
  left_join(total_dist, by = "year") %>%
  mutate(avg_elev = (start_elev + end_elev) / 2 ) %>%
  ggplot(aes(section_number, distance_km)) +
   geom_rect(aes(fill = dist_cyc_tot_km), 
                xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf, 
                alpha = 0.3) +
  geom_bar(stat = "identity", fill = "black") +
  theme(panel.background = element_rect(fill = "white")) +
  facet_wrap( ~ year, ncol = 7) +
  scale_fill_viridis_c(name = "Tour total distance cycled (km)") + 
  guides(fill = guide_colourbar(title.position = "right", 
                                    barwidth = 4, barheight = 40, 
                                    ticks = FALSE)) +
  theme(
    plot.title = element_text(
      size = 50,
      face = "bold",
      hjust = 0.5,
      margin = margin(t = 10, b = 40, unit = "pt")
    ),
    strip.text = element_text(size = 25, face = "bold", colour = "white"),
    axis.text.y = element_text(size = 25),
    axis.title.y = element_text(size = 40),
    axis.title.x = element_text(size = 40),
    axis.text.x = element_text(size = 0),
    legend.title = element_text(size = 40, angle = 90), 
    legend.text = element_text(size = 35, angle = 90),
    strip.background = element_rect(fill = "black"),
    axis.ticks.x = element_blank(), 
   ) +
  labs(title = "Distance cyclé dans les Tours de France", x = "Stage", y = "Distance cycled (km)")

ggsave("tour_distance.pdf", tour_dist, width = 1000, height = 1485, units = "mm", limitsize = FALSE)

```

# DATA SETUP FOR PLOTTING ROUTES

1) **Sections of the plotted route not cycled**
Create dataframe with coordinates of not cycled route of tour, ie space between end of one stage and start of the next. This will enable us to distinguish between 'cycled' stages and 'not cycled' stages in the plot using solid and dashed lines, respectively.
```{r not_cycled}
tour_not_cyc_stage <- tour_data %>%
  arrange(date) %>%
  #select(year, start_lat, start_long, end_lat, end_long) %>%
  group_by(year(date)) %>%
  mutate(end_prev_lat = if_else(date == min(date), 1E9, lag(end_lat)),
         end_prev_long = if_else(date == min(date), 1E9, lag(end_long))) %>%
  ungroup() %>%
  filter(end_prev_lat != 1E9) 

tour_not_cyc_stage <- tour_not_cyc_stage %>%
  select(year, end_prev_lat, end_prev_long, start_lat, start_long, terrain_group)
```

2) **Start and end of plotted route**
Add new column `tour_progress` to tour stages data to create colour gradient in chart.
```{r tour_progress}
tour_data <- tour_data %>%
  group_by(year) %>%
  filter(terrain_group != "Rest day") %>%
  mutate(tour_progress = stage_number / max(stage_number)) %>%
  ungroup()

filter(tour_data, is.na(tour_progress))
```

# PLOT TOUR ROUTES

Create `ggplot()` and write as function `plot_tour_ind()`.
```{r plot_tour_ind}
plot_tour_ind <- function(year_start, year_end, n_column) {
tour_subset_loc <- tour_all_loc %>%
  filter(terrain_group != "Rest day") %>%
  filter(year >= year_start & year <= year_end)

tour_subset_stage <- tour_data %>%
  filter(terrain_group != "Rest day") %>%
  filter(year >= year_start & year <= year_end)

tour_subset_not_cyc <- tour_not_cyc_stage %>%
  filter(terrain_group != "Rest day") %>%
  filter(year >= year_start & year <= year_end)

# Plot route for this year(s)
tours_plot <- borders_plot +
  # Plot start and end locations of tour
  geom_point(data = tour_subset_loc,
             aes(x = long, y = lat),
             size = 0) +
  # Plot cycled parts of tour as solid lines
  geom_segment(
    data = tour_subset_stage,
    aes(
      x = start_long,
      y = start_lat,
      xend = end_long,
      yend = end_lat,
      colour = tour_progress
    ),
   # arrow = arrow(length = unit(0.05, "npc"), angle = 20),
    size = 1.5
   ) +
  # Plot not cycled route between end of previous and start of next stage as dashed line
  geom_segment(
    data = tour_subset_not_cyc,
    aes(
      x = end_prev_long,
      y = end_prev_lat,
      xend = start_long,
      yend = start_lat
    ),
    size = 0.5,
    colour = "#2E2D2D",
    linetype = 3
  ) + 
  facet_wrap( ~ year, ncol = n_column)  

tours_plot
}
```

Plot individual years for investigation
```{r plot_ind}
plot_tour_ind(1947, 1960, n_column = 5)
```


Produce pdf of all Tour routes (tour_routes.pdf)
```{r plot_tours}
all_years_plot <- plot_tour_ind(1903, 2019, n_column = 10)

# Create vector with legend labels
leg_labels <- c("Start", "Finish")

all_years_plot <- all_years_plot +
  theme(plot.title = element_text(size = 50, face = "bold", hjust = 0.5, margin = margin(t = 10, b = 40, unit = "pt")),
        legend.position = "top",
        legend.title = element_text(size = 20, face = "bold", colour = "black"),
        legend.text = element_text(size = 20),
        legend.box.margin=margin(20,20,20,20),
        strip.text = element_text(size = 20, face = "bold")) +
  # Code to customise appearance of legend if using:
  scale_colour_gradient(name = "Tour progress", breaks = seq(0, 1, by = 1), limits = c(0, 1), labels = leg_labels, low = "green", high = "#01061d") +
  guides(colour = guide_legend(nrow = 1, keywidth = 0.1, keyheight = 0.4, default.unit = "inch")) +
  labs(title = "Les routes de Tour de France")

ggsave("tour_routes.pdf", all_years_plot, width = 1050, height = 1485, units = "mm", limitsize = FALSE)
```


